Application of Network Flow and Linear Programming to Scheduling of Clinicians

### 2019-07-30 paper/paper.methods.tex
Add section labels

In this section, we present an application of 0-1 Linear Programming to solve the clinician scheduling problem presented in Section \ref{???}. First, we describe the sets, indices and variables present in the formulation of the program. We then convert the constraints given in Table \ref{tbl:constraint-summary} into mathematical terms, and outline the objective function of the linear program.

\subsection{Sets and Indices}
We denote the set of all services (or sometimes referred to as divisions) %or I think services is good (vs. services/divisions) as you define two types of services (ID vs. HIV) in the problem statement
that clinicians in a single department can provide as $\mathcal{D}$. The following formulation assumes that all clinicians are able to provide all services. The set of all clinicians in the department is denoted as $\mathcal{C}$. The sets of blocks and weekends that clinicians will be assigned to are denoted as $\mathcal{B}$ and $\mathcal{W}$ respectively. The size of a block is not constrained and can be adapted %how?
to the needs of the given department. A subset of weekends are denoted $\mathcal{L}$, corresponding to established long/holiday weekends such as the Canadian Civic Day weekend in August. Lastly, we denote with $\mathcal{BR}_c$ and $\mathcal{WR}_c$ the block and weekend requests of clinicians as subsets of all blocks and weekends, respectively. For instance, if clinician $c$'s requests intersect with blocks 1 and 2, and weekend 1, then $\mathcal{BR}_c = \{1, 2\}$ and $\mathcal{WR}_c = \{1\}$. Table \ref{tbl:sets-indices} presents a summary of the sets and indices described.

@@

\subsection{Variables}
Since each clinician may be assigned to work on any service, during any block of the year, we denote such an assignment as $X_{c, b, d}$. A value of 1 indicates that the given clinician $c$ is assigned to cover division $d$ during block $b$. Similarly, we define weekend assignments as $Y_{c, w}$, although these are not indexed by service, as clinicians are expected to provide all services during the weekends. To optimize the soft constraint Block-Weekend Adjacency, we maximize the product $X_{c, b, d} \cdot Y_{c, w}$ for adjacent blocks and weekends. To formulate such an objective as a linear function of variables, we introduce another set of variables, denoted by $Z_{c, b, d}$, with additional constraints on its range. Further details regarding this variable are described in Sections \ref{???} and \ref{???}. We then introduce a set of constants $m_{c, d}$ and $M_{c, d}$ to constrain the minimal and maximal number of blocks each clinician is allowed to work during the year. Table \ref{tbl:variables-constants} presents a summary of the constants and variables in the problem.

@@

\subsection{Constraints}
We now present the hard constraints in Table \ref{tbl:constraint-summary} using the variables defined above.

@@

\subsection{Objectives}
As described in Section \ref{???}, the soft constraints of the clinician scheduling problem include: (1) satisfying clinician block off requests, (2) satisfying clinician weekend off requests, and (3) assigning weekends closer to blocks. We convert these soft constraints into linear objective functions of the binary variables defined in Section \ref{???}. Objectives (1) and (2) can be written as the following linear functions:  %minor suggestion - consider numbering the objectives for ease later when saying things like "these two objectives"...or writing the full name of the objectives instead of 'this' and 'these' as sometimes can get confusing what the 'this' and the 'these' refer to. 
\begin{align}
	&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{d=1}^{D} (-1)^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, d} \tag{Block Requests} \label{eqn:obj-block-requests}\\
	&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W} (-1)^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{Weekend Requests} \label{eqn:obj-weekend-requests}
\end{align}
where $\ind(\cdot)$ is the indicator function that has value 1 when the condition is met and 0 otherwise. In the above two objectives, we penalize any assignments that conflict with a block or weekend request, and aim to maximize the non-conflicting assignments. The Block-Weekend Adjacency is optimized by considering the product $X_{c, b, d}\cdot Y_{c, w}$ for values of $w$ ``adjacent'' to the value of $b$. For instance, clinicians might want to be assigned during a weekend that falls within an assigned block. So if clinician $c$ is assigned to work during block 3, corresponding to weeks 5 and 6 assuming 2-week blocks, they might also want to be assigned to work during weekend 5. In that case, we would like $X_{c, b=3, d} \cdot Y_{c, w=5}$ to be 1, since that indicates both variables are assigned. If at least one of the two variables is not assigned, the product will be 0. This leads to the maximization objective  %avoid editorial adjectives
\begin{align}
	&Q_3(X, Y) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{d=1}^{D} X_{c, b, d}\cdot Y_{c, w=\varphi(b)} \tag{Block-Weekend Adjacency} \label{eqn:obj-block-weekend-adj}
\end{align}
where $\varphi(b)$ maps a block one-to-one to an adjacent weekend, by some appropriate definition of adjacency. For instance, in the above example we will have $\varphi(b) = 2b - 1$. \\
% I really liked the example explanation!

>>>>

In this section, we present an application of 0-1 Linear Programming to solve the clinician scheduling problem presented in Section \ref{sec:problem}. First, we describe the sets, indices and variables present in the formulation of the program. We then convert the constraints given in Table \ref{tbl:constraint-summary} into mathematical terms, and outline the objective function of the linear program.

\subsection{Sets and Indices} \label{sec:meth-sets-indices}
We denote the set of all services (or sometimes referred to as divisions) %or I think services is good (vs. services/divisions) as you define two types of services (ID vs. HIV) in the problem statement
that clinicians in a single department can provide as $\mathcal{D}$. The following formulation assumes that all clinicians are able to provide all services. The set of all clinicians in the department is denoted as $\mathcal{C}$. The sets of blocks and weekends that clinicians will be assigned to are denoted as $\mathcal{B}$ and $\mathcal{W}$ respectively. The size of a block is not constrained and can be adapted %how?
to the needs of the given department. A subset of weekends are denoted $\mathcal{L}$, corresponding to established long/holiday weekends such as the Canadian Civic Day weekend in August. Lastly, we denote with $\mathcal{BR}_c$ and $\mathcal{WR}_c$ the block and weekend requests of clinicians as subsets of all blocks and weekends, respectively. For instance, if clinician $c$'s requests intersect with blocks 1 and 2, and weekend 1, then $\mathcal{BR}_c = \{1, 2\}$ and $\mathcal{WR}_c = \{1\}$. Table \ref{tbl:sets-indices} presents a summary of the sets and indices described.

@@

\subsection{Variables} \label{sec:meth-variables}
Since each clinician may be assigned to work on any service, during any block of the year, we denote such an assignment as $X_{c, b, d}$. A value of 1 indicates that the given clinician $c$ is assigned to cover division $d$ during block $b$. Similarly, we define weekend assignments as $Y_{c, w}$, although these are not indexed by service, as clinicians are expected to provide all services during the weekends. To optimize the soft constraint Block-Weekend Adjacency, we maximize the product $X_{c, b, d} \cdot Y_{c, w}$ for adjacent blocks and weekends. To formulate such an objective as a linear function of variables, we introduce another set of variables, denoted by $Z_{c, b, d}$, with additional constraints on its range. Further details regarding this variable are described in Section \ref{sec:meth-objectives}. We then introduce a set of constants $m_{c, d}$ and $M_{c, d}$ to constrain the minimal and maximal number of blocks each clinician is allowed to work during the year. Table \ref{tbl:variables-constants} presents a summary of the constants and variables in the problem.

@@

\subsection{Constraints} \label{sec:meth-constraints}
We now present the hard constraints in Table \ref{tbl:constraint-summary} using the variables defined above.

@@

\subsection{Objectives} \label{sec:meth-objectives}
As described in Section \ref{sec:problem}, the soft constraints of the clinician scheduling problem include: (1) satisfying clinician block off requests, (2) satisfying clinician weekend off requests, and (3) assigning weekends closer to blocks. We convert these soft constraints into linear objective functions of the binary variables defined in Section \ref{sec:meth-variables}. Objectives (1) and (2) can be written as the following linear functions:  %minor suggestion - consider numbering the objectives for ease later when saying things like "these two objectives"...or writing the full name of the objectives instead of 'this' and 'these' as sometimes can get confusing what the 'this' and the 'these' refer to. 
\begin{align}
	&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{d=1}^{D} (-1)^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, d} \tag{Block Requests} \label{eqn:obj-block-requests}\\
	&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W} (-1)^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{Weekend Requests} \label{eqn:obj-weekend-requests}
\end{align}
where $\ind(\cdot)$ is the indicator function that has value 1 when the condition is met and 0 otherwise. In the above two objectives, we penalize any assignments that conflict with a block or weekend request, and aim to maximize the non-conflicting assignments. The Block-Weekend Adjacency is optimized by considering the product $X_{c, b, d}\cdot Y_{c, w}$ for values of $w$ ``adjacent'' to the value of $b$. For instance, clinicians might want to be assigned during a weekend that falls within an assigned block. So if clinician $c$ is assigned to work during block 3, corresponding to weeks 5 and 6 assuming 2-week blocks, they might also want to be assigned to work during weekend 5. In that case, we would like $X_{c, b=3, d} \cdot Y_{c, w=5}$ to be 1, since that indicates both variables are assigned. If at least one of the two variables is not assigned, the product will be 0. This leads to the maximization objective  %avoid editorial adjectives
\begin{align}
	&Q_3(X, Y) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{d=1}^{D} X_{c, b, d}\cdot Y_{c, w=\varphi(b)} \tag{Block-Weekend Adjacency} \label{eqn:obj-block-weekend-adj}
\end{align}
where $\varphi(b)$ maps a block one-to-one to an adjacent weekend, by some appropriate definition of adjacency. For instance, in the above example we will have $\varphi(b) = 2b - 1$. \\
% I really liked the example explanation!

### 2019-07-30 paper/paper.results.tex
Add section labels

In order to evaluate the generated schedule and compare it with the manually created schedule, we outline the adherence of each schedule to the constraints presented in section \ref{???}. As we can see from table \ref{tbl:constraints-comparison}, the generated schedule was able to satisfy all mandatory constraints, however for the years of [???] it could not find an optimal schedule while ensuring that the block assignments are spread out ([???] constraint). On the other hand, we can see that the manually created schedule was not able to satisfy all mandatory constraints. In particular, we see that it contains consecutive block assignments for the years of [???], [...]. Evaluating the objectives, we see that [...\textit{request conflicts}]. Moreover, the manual schedule does not attempt to align weekend assignments with block assignments, unlike the optimal schedule found by the software.

@@

We plotted the effect of an increasing number of requests per clinician on the average runtime of the LP solver in figure \ref{fig:avg-runtime-requests}. We ran the program on a single-division department with 10 clinicians, and $R = \abs{\mathcal{WR}} + \abs{\mathcal{BR}} = \{1, 2, 3, 5, 10\}$ non-overlapping requests per clinician. The increase in the number of requests did not affect the runtime of the LP solver, indicating that it can handle and accommodate much flexibility in clinician requests.

@@

>>>>

In order to evaluate the generated schedule and compare it with the manually created schedule, we outline the adherence of each schedule to the constraints presented in table \ref{tbl:constraint-summary}. As we can see from table \ref{tbl:constraints-comparison}, the generated schedule was able to satisfy all mandatory constraints, however for the years of [???] it could not find an optimal schedule while ensuring that the block assignments are spread out ([???] constraint). On the other hand, we can see that the manually created schedule was not able to satisfy all mandatory constraints. In particular, we see that it contains consecutive block assignments for the years of [???], [...]. Evaluating the objectives, we see that [...\textit{request conflicts}]. Moreover, the manual schedule does not attempt to align weekend assignments with block assignments, unlike the optimal schedule found by the software.

@@

We plotted the effect of an increasing number of requests per clinician on the average runtime of the LP solver in figure \ref{fig:avg-runtime-requests}. We ran the program on a single-division department with 10 clinicians, and $R = \abs{\mathcal{WR}} + \abs{\mathcal{BR}} = \{1, 2, 3, 5, 10\}$ non-overlapping requests per clinician. The increase in the number of requests did not affect the runtime of the LP solver, indicating that it can handle and accommodate much flexibility in clinician requests. \\

@@

[...]
